---
title: "Protein Signatures as Biomarkers for Cardiovascular Disease in HIV Patients"
author: "Mathanki Mehra & Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

## Intro

Here we are analysing samples obtained before patients had a CVD event.
We have case samples from patients who later had a CVD event and a control groups
which in the follow up period did not have a CVD event.

## Load libraries

```{r,load-libraries}

library("limma")
library("mitch")

library("PLSDAbatch")

library("tibble")
library("tidyr")
library("readr")
library("dplyr")
library("ggplot2")
library("pheatmap")
library("janitor")
library("kableExtra")
library("eulerr")
library("beeswarm")

```

## Load files and clean the data tables

These files were generated by FragPipe.

```{r,load-files}

plexA <- read_tsv("abundance_protein_MD_plexA.tsv")
plexB <- read_tsv("abundance_protein_MD_plexB.tsv")

plexA <- janitor::clean_names(plexA)
plexB <- janitor::clean_names(plexB)

```

Extract out the protein quantifications.

```{r,keep-columns}

plexA <- dplyr::select(plexA, protein, gene,
                       x126, x127n, x127c, x128n, x128c,
                       x129n, x129c, x130n, x130c,
                       x131n, x131c, x132n)

plexB <- dplyr::select(plexB, protein, gene,
                       x126, x127n, x127c, x128n, x128c,
                       x129n, x129c, x130n, x130c,
                       x131n, x131c, x132n)
```

Rename TMT channels to sample IDs.

```{r,rename-channels}

annotationA <- c(
  x126="19191", x127n="34629", x127c="34862",
  x128n="40431", x128c="40475", x129n="35038",
  x129c="34866", x130n="41436", x130c="41383",
  x131n="40316", x131c="3746", x132n="102953"
)

annotationB <- c(
  x126="40797", x127n="4177", x127c="103376",
  x128n="47542", x128c="47442", x129n="42977",
  x129c="43081", x130n="113835", x130c="113856",
  x131n="19584", x131c="118117", x132n="118008"
)

colnames(plexA)[match(names(annotationA), colnames(plexA))] <- annotationA
colnames(plexB)[match(names(annotationB), colnames(plexB))] <- annotationB

```

Merge plexes.

```{r, merge-plexes}

merged <- full_join(plexA, plexB, by=c("protein","gene"))
head(merged) |>
    kbl(caption = "Example of merged data") |>
    kable_paper("hover", full_width = F)

dim(merged)

```

Expression matrix for Merged plexes including Plex A and Plex B.

```{r, create-matrix}

expr_combined <- as.data.frame(merged)
protein_info_combined <- expr_combined[, c("protein", "gene")]
rownames(expr_combined) <- expr_combined$protein
expr_combined <- expr_combined[, !(colnames(expr_combined) %in% c("protein","gene"))]
expr_combined <- as.matrix(expr_combined)
storage.mode(expr_combined) <- "numeric"

expr_A <- as.data.frame(plexA)
protein_info_A <- expr_A[, c("protein", "gene")]
rownames(expr_A) <- expr_A$protein
expr_A <- expr_A[, !(colnames(expr_A) %in% c("protein","gene"))]
expr_A <- as.matrix(expr_A)
storage.mode(expr_A) <- "numeric"
head(expr_A) |>
    kbl(caption = "Example of Plex A data") |>
    kable_paper("hover", full_width = F)


dim(expr_A)

expr_B <- as.data.frame(plexB)
protein_info_B <- expr_B[, c("protein", "gene")]
rownames(expr_B) <- expr_B$protein
expr_B <- expr_B[, !(colnames(expr_B) %in% c("protein","gene"))]
expr_B <- as.matrix(expr_B)
storage.mode(expr_B) <- "numeric"
head(expr_B)|>
    kbl(caption = "Example of Plex B data") |>
    kable_paper("hover", full_width = F)

dim(expr_B)

```

## Filter proteins

Keep proteins which are present in >=50% of samples.

```{r, filter1}

keep_combined <- rowSums(!is.na(expr_combined)) >= ncol(expr_combined)*0.5
expr_combined_filt <- expr_combined[keep_combined, ]
dim(expr_combined_filt)

keep_A <- rowSums(!is.na(expr_A)) >= ncol(expr_A)*0.5
expr_A_filt <- expr_A[keep_A, ]
dim(expr_A_filt)

keep_B <- rowSums(!is.na(expr_B)) >= ncol(expr_B)*0.5
expr_B_filt <- expr_B[keep_B, ]
dim(expr_B_filt)

```

## QC before normalization

```{r, qc-pre-normalization}

qc_sample_combined <- data.frame(
  Sample = colnames(expr_combined_filt),
  Detected_Proteins = colSums(!is.na(expr_combined_filt)),
  Missing_Percent = colMeans(is.na(expr_combined_filt)) * 100,
  Median_Intensity = apply(expr_combined_filt, 2, median, na.rm = TRUE),
  SD_Intensity = apply(expr_combined_filt, 2, sd, na.rm = TRUE)
)

# Assign plex
plexA_samples <- c(
  "19191","34629","34862","40431","40475","35038",
  "34866","41436","41383","40316","3746","102953"
)

qc_sample_combined$Plex <- ifelse(qc_sample_combined$Sample %in% plexA_samples, "A", "B")

# QC summary by plex
qc_plex_combined <- qc_sample_combined %>%
  group_by(Plex) %>%
  summarise(
    N_Samples = n(),
    Mean_Detected_Proteins = mean(Detected_Proteins),
    Mean_Missing_Percent = mean(Missing_Percent),
    Mean_Median_Intensity = mean(Median_Intensity),
    .groups = "drop"
  )

qc_sample_combined |>
    kbl(caption = "QC on combined data") |>
    kable_paper("hover", full_width = F)

qc_plex_combined |>
    kbl(caption = "Summary of Plex data QC") |>
    kable_paper("hover", full_width = F)

```

Now we have some QC plots.
The first one is the missingness plot.

```{r, missingnessplot}

qc_long_combined <- as.data.frame(expr_combined_filt) %>%
  rownames_to_column("Protein") %>%
  pivot_longer(-Protein, names_to = "Sample", values_to = "Intensity") %>%
  filter(is.finite(Intensity))

ggplot(qc_sample_combined,
       aes(x = reorder(Sample, Missing_Percent),
           y = Missing_Percent,
           fill = Plex)) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  labs(title = "Missingness per Sample",
       x = "Sample", y = "% Missing")


```

Detected proteins plot.

```{r, detected-proteins-plot}

ggplot(qc_sample_combined,
       aes(x = reorder(Sample, Detected_Proteins),
           y = Detected_Proteins,
           fill = Plex)) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  labs(title = "Detected Proteins per Sample",
       x = "Sample", y = "Detected Proteins")

```

Density plot.

```{r, density-plot}

ggplot(qc_long_combined,
       aes(x = Intensity, color = Sample)) +
  geom_density(linewidth = 0.8) +
  theme_bw() +
  labs(title = "Intensity Distribution per Sample",
       x = "Intensity", y = "Density")

```

Plex-level comparison.

```{r, plex-level-comparison}

ggplot(qc_plex_combined,
       aes(x = Plex, y = Mean_Detected_Proteins, fill = Plex)) +
  geom_col() +
  theme_bw() +
  labs(title = "Mean Detected Proteins by Plex",
       x = "Plex", y = "Mean Detected Proteins")

```

## Correlation heatmap

A correlation heatmap shows the similarity between samples.

It shows that the samples are clustered into clusters based on plex A/B.

```{r,corr_heatmap1, width=7, height=7}

cor_pre_combined <- cor(expr_combined_filt, use = "pairwise.complete.obs")

pheatmap(cor_pre_combined,
         main = "Sample Correlation Heatmap (Pre-normalization)")

```

## Quantile normalization and log2 transform

This needs to be run for combined dataset and separate plexes as well.

```{r, quantile-normalisation-log-tfm}

expr_combined_norm <- normalizeBetweenArrays(expr_combined_filt, method = "quantile")
expr_combined_log <- log2(expr_combined_norm + 1)
boxplot(expr_combined_log, main = "Post-normalization for both plexes", las = 2)

# Plex A
expr_A_norm <- normalizeBetweenArrays(expr_A_filt, method = "quantile")
expr_A_log <- log2(expr_A_norm + 1)
boxplot(expr_A_log, main = "Post-normalization for Plex A", las = 2)

# plex B
expr_B_norm <- normalizeBetweenArrays(expr_B_filt, method = "quantile")
expr_B_log <- log2(expr_B_norm + 1)
boxplot(expr_B_log, main = "Post-normalization for Plex B", las = 2)

```

## QC after normalization

After normalisation, the signal distributions are uniform, but there's a shoulder.

The correlation structure is retained as expected.

```{r, qc-post-normalization}

qc_long_norm_combined <- as.data.frame(expr_combined_log) %>%
  rownames_to_column("Protein") %>%
  pivot_longer(-Protein, names_to = "Sample", values_to = "Intensity") %>%
  filter(is.finite(Intensity))

ggplot(qc_long_norm_combined,
       aes(x = Intensity, color = Sample)) +
  geom_density(linewidth = 0.8) +
  theme_bw() +
  labs(title = "Post-normalization Intensity Distribution (Combined)",
       x = "Log2 Intensity",
       y = "Density")


cor_post_combined <- cor(expr_combined_log, use = "pairwise.complete.obs")

pheatmap(cor_post_combined,
         main = "Sample Correlation Heatmap (Post-normalization)")

```

## Impute missing values

Here we are using the median method for imputation of missing values.

```{r, impute1}

# Combine Plexes
expr_combined_imp <- expr_combined_log
for(i in 1:ncol(expr_combined_imp)){
  expr_combined_imp[is.na(expr_combined_imp[,i]), i] <-
    median(expr_combined_imp[,i], na.rm = TRUE)
}

# Plex A
expr_A_imp <- expr_A_log

for(i in 1:ncol(expr_A_imp)){
  expr_A_imp[is.na(expr_A_imp[,i]), i] <-
    median(expr_A_imp[,i], na.rm = TRUE)
}

# Plex B
expr_B_imp <- expr_B_log

for(i in 1:ncol(expr_B_imp)){
  expr_B_imp[is.na(expr_B_imp[,i]), i] <-
    median(expr_B_imp[,i], na.rm = TRUE)
}
sum(is.na(expr_combined_imp))
sum(is.na(expr_A_imp))
sum(is.na(expr_B_imp))

```

Next, we will removing Zero Variance Proteins because they cannot be statistically significant and may
cause other problems later on.

```{r,remove-zero-var}

expr_combined_imp <- expr_combined_imp[apply(expr_combined_imp, 1, var) != 0, ]
expr_A_imp <- expr_A_imp[apply(expr_A_imp, 1, var) != 0, ]
expr_B_imp <- expr_B_imp[apply(expr_B_imp, 1, var) != 0, ]

```

## Metadata Creation

```{r,metadata}

# Create combined metadata from expression matrix
meta_combined <- data.frame(
  Sample = colnames(expr_combined_imp)
)

# Define Plex (based on known Plex A samples)

plexA_samples <- c(
  "19191","34629","34862","40431","40475","35038",
  "34866","41436","41383","40316","3746","102953"
)

meta_combined$Plex <- ifelse(meta_combined$Sample %in% plexA_samples,
                             "A", "B")

meta_combined$Plex <- factor(meta_combined$Plex)

# Define biological group (Case vs Control)

case_samples <- c(
  "19191","40797","34862","103376","40475","47442",
  "34866","43081","41383","113856","3746","118117"
)

meta_combined$Group <- ifelse(meta_combined$Sample %in% case_samples,
                              "Case", "Control")

meta_combined$Group <- factor(meta_combined$Group,
                              levels = c("Control","Case"))

# Final formatting

rownames(meta_combined) <- meta_combined$Sample

# Create Plex-specific metadata (CORRECT WAY)
meta_A <- meta_combined[meta_combined$Plex == "A", ]
meta_B <- meta_combined[meta_combined$Plex == "B", ]

# Align with expression matrices (CRITICAL)
meta_A <- meta_A[colnames(expr_A_imp), ]
meta_B <- meta_B[colnames(expr_B_imp), ]

# Keep only needed columns
meta_A <- meta_A[, c("Sample","Group")]
meta_B <- meta_B[, c("Sample","Group")]

# SANITY CHECKS (DO NOT SKIP)
# Alignment
all(rownames(meta_A) == colnames(expr_A_imp))
all(rownames(meta_B) == colnames(expr_B_imp))
all(rownames(meta_combined) == colnames(expr_combined_imp))

# Distribution checks
table(meta_combined$Plex, meta_combined$Group)
table(meta_A$Group)
table(meta_B$Group)

```

## PCA Before Batch Correction for combined plexes

```{r, pca1}

pca_pre <- prcomp(t(expr_combined_imp), center = TRUE, scale. = TRUE)

pca_pre_df <- data.frame(
  PC1 = pca_pre$x[,1],
  PC2 = pca_pre$x[,2],
  Sample = colnames(expr_combined_imp),
  Plex = meta_combined$Plex,
  Group = meta_combined$Group
)

ggplot(pca_pre_df, aes(x = PC1, y = PC2, color = Plex, shape = Group)) +
  geom_point(size = 4) +
  geom_text(aes(label = Sample), vjust = -0.5, size = 3) +
  theme_bw() +
  labs(title = "PCA Before Batch Correction",
       x = "PC1",
       y = "PC2")

```

## Batch correction

```{r, batch correction}

expr_combined_corrected <- removeBatchEffect(
  expr_combined_imp,
  batch = meta_combined$Plex,
  design = model.matrix(~ Group, data = meta_combined)
)

# Quick check
head(expr_combined_corrected[,1:5])
# Dimensions unchanged
dim(expr_combined_corrected)

```

## PCA On Combined Plexes after removing Batch effect

```{r, pca2}

pca_post <- prcomp(t(expr_combined_corrected), center = TRUE, scale. = TRUE)

pca_post_df <- data.frame(
  PC1 = pca_post$x[,1],
  PC2 = pca_post$x[,2],
  Sample = colnames(expr_combined_corrected),
  Plex = meta_combined$Plex,
  Group = meta_combined$Group
)

ggplot(pca_post_df, aes(x = PC1, y = PC2, color = Group, shape = Plex)) +
  geom_point(size = 4) +
  geom_text(aes(label = Sample), vjust = -0.5, size = 3) +
  theme_bw() +
  labs(title = "PCA After Batch Correction",
       x = "PC1",
       y = "PC2")
```

## Differential Expression Analysis

### Combined plexes

First analysis includes batch correction at the limma step

```{r, de_combined1}

design_combined <- model.matrix(~ Plex + Group, data = meta_combined)

fit_combined <- lmFit(expr_combined_imp, design_combined)
fit_combined <- eBayes(fit_combined)
results_combined <- topTable(
  fit_combined,
  coef = "GroupCase",
  number = Inf,
  sort.by = "P"
)
results_combined$Protein <- rownames(results_combined)
# Add gene annotation
results_combined <- merge(
  results_combined,
  protein_info_combined,
  by.x = "Protein",
  by.y = "protein",
  all.x = TRUE
)

results_combined <- results_combined[order(results_combined$P.Value),]

head(results_combined) |>
    kbl(caption = "Top ranked proteins by p-value") |>
    kable_paper("hover", full_width = F)

```

Batch correction was done separately before limma.

```{r, de_combined2}

design_combined <- model.matrix(~ Group, data = meta_combined)

fit_combined <- lmFit(expr_combined_corrected, design_combined)
fit_combined <- eBayes(fit_combined)
results_combined <- topTable(
  fit_combined,
  coef = "GroupCase",
  number = Inf,
  sort.by = "P"
)
results_combined$Protein <- rownames(results_combined)
# Add gene annotation
results_combined <- merge(
  results_combined,
  protein_info_combined,
  by.x = "Protein",
  by.y = "protein",
  all.x = TRUE
)

results_combined <- results_combined[order(results_combined$P.Value),]

head(results_combined) |>
    kbl(caption = "Top ranked proteins by p-value") |>
    kable_paper("hover", full_width = F)

```

Make sure to do volcano plot, box/beeswarm plot and heatmap.

Volcano plot

```{r,volcano,fig.height=5,fig.width=5}

sig <- subset(results_combined,adj.P.Val<0.05)

plot(results_combined$logFC, -log10(results_combined$P.Value),
  cex=0.6,pch=19,col="darkgray",
  xlab="log2 fold change",ylab="-log10(p-value)")

abline(v=0,lty=2,lwd=1)
points(sig$logFC, -log10(sig$P.Value) , col="red",pch=19,cex=0.7)
text(sig$logFC+0.015, -log10(sig$P.Value), labels=sig$gene)

```

```{r,maplot,fig.height=5,fig.width=5}

sig <- subset(results_combined,adj.P.Val<0.05)

plot(log10(results_combined$AveExpr),results_combined$logFC,
  cex=0.6,pch=19,col="darkgray",
  xlab="log10 basemean",ylab="log2 fold change")

abline(h=0,lty=2,lwd=1)
points(log10(sig$AveExpr),sig$logFC,col="red",pch=19,cex=0.7)
text(log10(sig$AveExpr)+0.004,sig$logFC,labels=sig$gene)

```

Beeswarm plot using batch corrected data.

```{r,beeswarm,fig.width=9,fig.height=9}

par(mfrow=c(3,3))

null <- lapply(1:9,function(i) {
  mygene <- results_combined$gene[i]
  myprot <- results_combined$Protein[i]
  mydat <- expr_combined_corrected[which(rownames(expr_combined_corrected) == myprot),]
  mygroups <- meta_combined[match(names(mydat),meta_combined$Sample),3]
  ctrl <- mydat[which(mygroups=="Control")]
  case <- mydat[which(mygroups=="Case")]
  xl <- list("Ctrl"=ctrl,"Case"=case)
  boxplot(xl,ylab="Relative protein expression",main=mygene,col="white",cex=0)
  beeswarm(xl,add=TRUE,cex=1.5,pch=19,col="darkgray")
} )

```

Now look at the ratio between MAPK1 and INF2.

```{r,ratio1}

prota="sp|P28482|MK01_HUMAN"
protb="sp|Q27J81|INF2_HUMAN"
dat_a <- expr_combined_corrected[which(rownames(expr_combined_corrected)==prota),]
dat_b <- expr_combined_corrected[which(rownames(expr_combined_corrected)==protb),]
abrat <- dat_a / dat_b
case_samples <- meta_combined[which(meta_combined$Group=="Case"),"Sample"]
ctrl_samples <- meta_combined[which(meta_combined$Group=="Control"),"Sample"]

case_dat <- abrat[which(names(abrat) %in% case_samples)]
ctrl_dat <- abrat[which(names(abrat) %in% ctrl_samples)]

rl <- list("Ctrl"=ctrl_dat, "Case"=case_dat)
boxplot(rl,ylab="Ratio",main="Ratio MAPK1/INF2",col="white",cex=0)
beeswarm(rl,add=TRUE,cex=1.5,pch=19,col="darkgray")
t.test(case_dat,ctrl_dat)

```

## Separate plexes

Analyse plex A and plex B.

```{r, de_plex_A}

design_A <- model.matrix(~ Group, data = meta_A)
fit_A <- lmFit(expr_A_imp, design_A)
fit_A <- eBayes(fit_A)
results_A <- topTable(
  fit_A,
  coef = "GroupCase",
  number = Inf,
  sort.by = "P"
)
results_A$Protein <- rownames(results_A)
# Add gene annotation
results_A <- merge(
  results_A,
  protein_info_A,
  by.x = "Protein",
  by.y = "protein",
  all.x = TRUE
)

# View
results_A <- results_A[order(results_A$P.Value),]

head(results_A,10) |>
    kbl(caption = "Top ranked proteins by p-value in plex A") |>
    kable_paper("hover", full_width = F)

```

Plex B.

```{r,de_plex_B}

design_B <- model.matrix(~ Group, data = meta_B)
fit_B <- lmFit(expr_B_imp, design_B)
fit_B <- eBayes(fit_B)
results_B <- topTable(
  fit_B,
  coef = "GroupCase",
  number = Inf,
  sort.by = "P"
)
results_B$Protein <- rownames(results_B)
# Add gene annotation
results_B <- merge(
  results_B,
  protein_info_B,
  by.x = "Protein",
  by.y = "protein",
  all.x = TRUE
)

# View
results_B <- results_B[order(results_B$P.Value),]

head(results_B,10) |>
    kbl(caption = "Top ranked proteins by p-value in plex B") |>
    kable_paper("hover", full_width = F)

```

## Comparison of Plex A and B results

The top 50 up and down regulated proteins are selected from Plex A and B and these lists were compared.


```{r, compare_A_B}

plexA_up <- head(subset(results_A,logFC>0),50)[,1]
plexA_dn <- head(subset(results_A,logFC<0),50)[,1]
plexB_up <- head(subset(results_B,logFC>0),50)[,1]
plexB_dn <- head(subset(results_B,logFC<0),50)[,1]

v1 <- list("A up"=plexA_up, "A dn"=plexA_dn, "B up"=plexB_up, "B dn"=plexB_dn)

plot(euler(v1), quantities = TRUE)

intersect(plexA_up,plexB_up)

intersect(plexA_dn,plexB_dn)

intersect(plexA_up,plexB_dn)

intersect(plexB_up,plexA_dn)

```

## Ranking of proteins by logFC

```{r, rank_analysis}

results_A$score <- sign(results_A$logFC) * -log10(results_A$P.Value)
results_B$score <- sign(results_B$logFC) * -log10(results_B$P.Value)

results_m <- merge(results_A,results_B,by="Protein")

ranks_m <- results_m[,c("Protein","score.x","score.y")]
rownames(ranks_m) <- ranks_m$Protein ; ranks_m$Protein=NULL
colnames(ranks_m) <- c("Plex A","Plex B")

plot(ranks_m)

mylm <- lm(ranks_m$`Plex B` ~ ranks_m$`Plex A`)
abline(mylm,lwd=2,lty=2,col="red")
COR=signif(cor(ranks_m),3)[2,1]
COR
mtext(paste("Pearson correlation=",COR))

```

## Cross batch analysis

Look at the proteins that are differentially expressed across batches.

IGHV and IHKV are over-represented.

Would be interesting to look at the composition of these proteins to understand the batch effect better.

```{r,crossbatch}

design_batch <- model.matrix(~ Plex , data = meta_combined)

fit_batch <- lmFit(expr_combined_imp, design_batch)

fit_batch <- eBayes(fit_batch)

summary(decideTests(fit_batch))

results_batch <- topTable(
  fit_batch,
  coef = ncol(summary(decideTests(fit_batch))),
  number = Inf,
  sort.by = "P"
)

results_batch$Protein <- rownames(results_batch)

# Add gene annotation
results_batch <- merge(
  results_batch,
  protein_info_combined,
  by.x = "Protein",
  by.y = "protein",
  all.x = TRUE
)

results_batch <- results_batch[order(results_batch$P.Value),]

head(results_batch,10) |>
    kbl(caption = "Proteins related to Plex A/B differences.") |>
    kable_paper("hover", full_width = F)

head(subset(results_batch,logFC>0),20) |>
    kbl(caption = "Proteins with higher expression in Plex B.") |>
    kable_paper("hover", full_width = F)

head(subset(results_batch,logFC<0),20) |>
    kbl(caption = "Proteins with higher expression in Plex A.") |>
    kable_paper("hover", full_width = F)

```

## PLSDA-batch

```{r,plsda-batch}

expr_combined_pl_corr <- PLSDA_batch(
    X = t(expr_combined_imp),
    Y.trt = meta_combined$Group, Y.bat = meta_combined$Plex,
    ncomp.trt = 1, ncomp.bat = 1,
    mode = "regression"
)

expr_combined_pl_corr$X.nobatch[1:5,1:7]

expr_combined_pl_corrmx <- t(expr_combined_pl_corr$X.nobatch)
expr_combined_pl_corrmx[1:6,1:4]

```

## Limma on PLSDA-batch corrected data

```{r,plsda-batch-limma}

# stash the old analysis
results_combined0 <- results_combined

design_combined <- model.matrix(~ Group, data = meta_combined)

fit_combined <- lmFit(expr_combined_pl_corrmx, design_combined)
fit_combined <- eBayes(fit_combined)
results_combined <- topTable(
  fit_combined,
  coef = "GroupCase",
  number = Inf,
  sort.by = "P"
)
results_combined$Protein <- rownames(results_combined)
# Add gene annotation
results_combined <- merge(
  results_combined,
  protein_info_combined,
  by.x = "Protein",
  by.y = "protein",
  all.x = TRUE
)

results_combined <- results_combined[order(results_combined$P.Value),]

head(results_combined) |>
    kbl(caption = "Top ranked proteins by p-value (limma done on PLSDA-batch-corrected data)") |>
    kable_paper("hover", full_width = F)

```

Beeswarm plot using batch corrected data.

```{r,beeswarm2,fig.width=9,fig.height=9}

par(mfrow=c(3,3))

null <- lapply(1:9,function(i) {
  mygene <- results_combined$gene[i]
  myprot <- results_combined$Protein[i]
  mydat <- expr_combined_pl_corrmx[which(rownames(expr_combined_pl_corrmx) == myprot),]
  mygroups <- meta_combined[match(names(mydat),meta_combined$Sample),3]
  ctrl <- mydat[which(mygroups=="Control")]
  case <- mydat[which(mygroups=="Case")]
  xl <- list("Ctrl"=ctrl,"Case"=case)
  boxplot(xl,ylab="Relative protein expression",main=mygene,col="white",cex=0)
  beeswarm(xl,add=TRUE,cex=1.5,pch=19,col="darkgray")
} )

```

Now look at the ratio between MAPK1 and INF2.

```{r,ratio2}

prota="sp|P28482|MK01_HUMAN"
protb="sp|Q27J81|INF2_HUMAN"
dat_a <- expr_combined_pl_corrmx[which(rownames(expr_combined_pl_corrmx)==prota),]
dat_b <- expr_combined_pl_corrmx[which(rownames(expr_combined_pl_corrmx)==protb),]
abrat <- dat_a / dat_b
case_samples <- meta_combined[which(meta_combined$Group=="Case"),"Sample"]
ctrl_samples <- meta_combined[which(meta_combined$Group=="Control"),"Sample"]

case_dat <- abrat[which(names(abrat) %in% case_samples)]
ctrl_dat <- abrat[which(names(abrat) %in% ctrl_samples)]

rl <- list("Ctrl"=ctrl_dat, "Case"=case_dat)
boxplot(rl,ylab="Ratio",main="Ratio MAPK1/INF2",col="white",cex=0)
beeswarm(rl,add=TRUE,cex=1.5,pch=19,col="darkgray")
t.test(case_dat,ctrl_dat)

```

## Incorporate time-to-event

```{r,time-to-event}

tte <- read.table("time-to-event.tsv",header=TRUE)

meta_combined$timetoevent <- tte[match(meta_combined$Sample,tte$SampleID),"time_to_event"]

```

## Pathway Enrichment Analysis

Using the combined protein profiles.

Some of the pathways are known to be involved in CVD.

Can try GOBP as well.

```{r,mitch1}

m <- mitch_import(results_combined,DEtype="limma",geneIDcol="gene")

reactome_sets <- gmt_import("ReactomePathways_2026-04-02.gmt")

mres <- mitch_calc(m,  reactome_sets, minsetsize = 5, cores=8)

head(mres$enrichment_result,10) |>
    kbl(caption = "Top ranked Reactome pathways by p-value") |>
    kable_paper("hover", full_width = F)

head(subset(mres$enrichment_result,s.dist>0),10) |>
    kbl(caption = "Top ranked upregulated Reactome pathways by p-value") |>
    kable_paper("hover", full_width = F)

head(subset(mres$enrichment_result,s.dist<0),10) |>
    kbl(caption = "Top ranked upregulated Reactome pathways by p-value") |>
    kable_paper("hover", full_width = F)

if ( file.exists("mitchreport1.html") ) { unlink("mitchreport1.html") }
mitch_report(mres,"mitchreport1.html")

```

GOBP

```{r,mitch2}

#download.file("https://data.broadinstitute.org/gsea-msigdb/msigdb/release/2026.1.Hs/c5.go.v2026.1.Hs.symbols.gmt",destfile="  c5.go.v2026.1.Hs.symbols.gmt")
#https://data.broadinstitute.org/gsea-msigdb/msigdb/release/2026.1.Hs/c5.go.bp.v2026.1.Hs.symbols.gmt
gobp_sets <- gmt_import("c5.go.bp.v2026.1.Hs.symbols.gmt")

names(gobp_sets) <- gsub("GOBP_","",names(gobp_sets))

names(gobp_sets) <- gsub("_"," ",names(gobp_sets))

mres2 <- mitch_calc(m,  gobp_sets, minsetsize = 5, cores=8)

head(mres2$enrichment_result,10) |>
    kbl(caption = "Top ranked GO sets by p-value") |>
    kable_paper("hover", full_width = F)

head(subset(mres2$enrichment_result,s.dist>0),10) |>
    kbl(caption = "Top ranked upregulated GO sets by p-value") |>
    kable_paper("hover", full_width = F)

head(subset(mres2$enrichment_result,s.dist<0),10) |>
    kbl(caption = "Top ranked upregulated GO sets by p-value") |>
    kable_paper("hover", full_width = F)

if ( file.exists("mitchreport2.html") ) { unlink("mitchreport2.html") }
mitch_report(mres2,"mitchreport2.html")

```

## Session Info

```{r,session}

sessionInfo()

save.image("analysis4_mz.Rdata")

```

