---
title: "Protein Signatures as Biomarkers for Cardiovascular Disease in HIV Patients"
author: "Mathanki Mehra & Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

## Intro

Here we are analysing samples obtained before patients had a CVD event.
We have case samples from patients who later had a CVD event and a control groups
which in the follow up period did not have a CVD event.

## Load libraries

```{r,load-libraries}

library(BiocManager)
library(dplyr)
library(tidyverse)
library(readr)
library(janitor)
library(limma)
library(pheatmap)
library(clusterProfiler)
library(org.Hs.eg.db)
library(ggplot2)
library(umap)
library(ReactomePA)
library(tibble)
library(knitr)
library(mitch)
library(stringr)
library(msigdbr)
library(beeswarm)
library("kableExtra")
library("eulerr")

```

## Load files and clean the data tables

These files were generated by FragPipe.

```{r,load-files}

plexA <- read_tsv("abundance_protein_MD_plexA.tsv")
plexB <- read_tsv("abundance_protein_MD_plexB.tsv")

plexA <- janitor::clean_names(plexA)
plexB <- janitor::clean_names(plexB)

```

Extract out the protein quantifications.

```{r,keep-columns}

plexA <- dplyr::select(plexA, protein, gene,
                       x126, x127n, x127c, x128n, x128c,
                       x129n, x129c, x130n, x130c,
                       x131n, x131c, x132n)

plexB <- dplyr::select(plexB, protein, gene,
                       x126, x127n, x127c, x128n, x128c,
                       x129n, x129c, x130n, x130c,
                       x131n, x131c, x132n)
```

Rename TMT channels to sample IDs.

```{r,rename-channels}

annotationA <- c(
  x126="19191", x127n="34629", x127c="34862",
  x128n="40431", x128c="40475", x129n="35038",
  x129c="34866", x130n="41436", x130c="41383",
  x131n="40316", x131c="3746", x132n="102953"
)

annotationB <- c(
  x126="40797", x127n="4177", x127c="103376",
  x128n="47542", x128c="47442", x129n="42977",
  x129c="43081", x130n="113835", x130c="113856",
  x131n="19584", x131c="118117", x132n="118008"
)

colnames(plexA)[match(names(annotationA), colnames(plexA))] <- annotationA
colnames(plexB)[match(names(annotationB), colnames(plexB))] <- annotationB

```

Merge plexes.

```{r, merge-plexes}

merged <- full_join(plexA, plexB, by=c("protein","gene"))
head(merged) |>
    kbl(caption = "Example of merged data") |>
    kable_paper("hover", full_width = F)

dim(merged)

```

Expression matrix for Merged plexes including Plex A and Plex B.

```{r, create-matrix}

expr_combined <- as.data.frame(merged)
protein_info_combined <- expr_combined[, c("protein", "gene")]
rownames(expr_combined) <- expr_combined$protein
expr_combined <- expr_combined[, !(colnames(expr_combined) %in% c("protein","gene"))]
expr_combined <- as.matrix(expr_combined)
storage.mode(expr_combined) <- "numeric"

expr_A <- as.data.frame(plexA)
protein_info_A <- expr_A[, c("protein", "gene")]
rownames(expr_A) <- expr_A$protein
expr_A <- expr_A[, !(colnames(expr_A) %in% c("protein","gene"))]
expr_A <- as.matrix(expr_A)
storage.mode(expr_A) <- "numeric"
head(expr_A) |>
    kbl(caption = "Example of Plex A data") |>
    kable_paper("hover", full_width = F)


dim(expr_A)

expr_B <- as.data.frame(plexB)
protein_info_B <- expr_B[, c("protein", "gene")]
rownames(expr_B) <- expr_B$protein
expr_B <- expr_B[, !(colnames(expr_B) %in% c("protein","gene"))]
expr_B <- as.matrix(expr_B)
storage.mode(expr_B) <- "numeric"
head(expr_B)|>
    kbl(caption = "Example of Plex B data") |>
    kable_paper("hover", full_width = F)

dim(expr_B)

```

## Filter proteins

Keep proteins which are present in >=50% of samples.

```{r, filter1}

keep_combined <- rowSums(!is.na(expr_combined)) >= ncol(expr_combined)*0.5
expr_combined_filt <- expr_combined[keep_combined, ]
dim(expr_combined_filt)

keep_A <- rowSums(!is.na(expr_A)) >= ncol(expr_A)*0.5
expr_A_filt <- expr_A[keep_A, ]
dim(expr_A_filt)

keep_B <- rowSums(!is.na(expr_B)) >= ncol(expr_B)*0.5
expr_B_filt <- expr_B[keep_B, ]
dim(expr_B_filt)

```

## QC before normalization

```{r, qc-pre-normalization}

qc_sample_combined <- data.frame(
  Sample = colnames(expr_combined_filt),
  Detected_Proteins = colSums(!is.na(expr_combined_filt)),
  Missing_Percent = colMeans(is.na(expr_combined_filt)) * 100,
  Median_Intensity = apply(expr_combined_filt, 2, median, na.rm = TRUE),
  SD_Intensity = apply(expr_combined_filt, 2, sd, na.rm = TRUE)
)

# Assign plex
plexA_samples <- c(
  "19191","34629","34862","40431","40475","35038",
  "34866","41436","41383","40316","3746","102953"
)

qc_sample_combined$Plex <- ifelse(qc_sample_combined$Sample %in% plexA_samples, "A", "B")

# QC summary by plex
qc_plex_combined <- qc_sample_combined %>%
  group_by(Plex) %>%
  summarise(
    N_Samples = n(),
    Mean_Detected_Proteins = mean(Detected_Proteins),
    Mean_Missing_Percent = mean(Missing_Percent),
    Mean_Median_Intensity = mean(Median_Intensity),
    .groups = "drop"
  )

qc_sample_combined |>
    kbl(caption = "QC on combined data") |>
    kable_paper("hover", full_width = F)

qc_plex_combined |>
    kbl(caption = "Summary of Plex data QC") |>
    kable_paper("hover", full_width = F)

```

Now we have some QC plots.
The first one is the missingness plot.

```{r, missingnessplot}

qc_long_combined <- as.data.frame(expr_combined_filt) %>%
  rownames_to_column("Protein") %>%
  pivot_longer(-Protein, names_to = "Sample", values_to = "Intensity") %>%
  filter(is.finite(Intensity))

ggplot(qc_sample_combined,
       aes(x = reorder(Sample, Missing_Percent),
           y = Missing_Percent,
           fill = Plex)) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  labs(title = "Missingness per Sample",
       x = "Sample", y = "% Missing")


```

Detected proteins plot.

```{r, detected-proteins-plot}

ggplot(qc_sample_combined,
       aes(x = reorder(Sample, Detected_Proteins),
           y = Detected_Proteins,
           fill = Plex)) +
  geom_col() +
  coord_flip() +
  theme_bw() +
  labs(title = "Detected Proteins per Sample",
       x = "Sample", y = "Detected Proteins")

```

Density plot.

```{r, density-plot}

ggplot(qc_long_combined,
       aes(x = Intensity, color = Sample)) +
  geom_density(linewidth = 0.8) +
  theme_bw() +
  labs(title = "Intensity Distribution per Sample",
       x = "Intensity", y = "Density")

```

Plex-level comparison.

```{r, plex-level-comparison}

ggplot(qc_plex_combined,
       aes(x = Plex, y = Mean_Detected_Proteins, fill = Plex)) +
  geom_col() +
  theme_bw() +
  labs(title = "Mean Detected Proteins by Plex",
       x = "Plex", y = "Mean Detected Proteins")

```

## Correlation heatmap

A correlation heatmap shows the similarity between samples.

It shows that the samples are clustered into clusters based on plex A/B.

```{r,corr_heatmap1, width=7, height=7}

cor_pre_combined <- cor(expr_combined_filt, use = "pairwise.complete.obs")

pheatmap(cor_pre_combined,
         main = "Sample Correlation Heatmap (Pre-normalization)")

```

## Quantile normalization and log2 transform

This needs to be run for combined dataset and separate plexes as well.

```{r, quantile-normalisation-log-tfm}

expr_combined_norm <- normalizeBetweenArrays(expr_combined_filt, method = "quantile")
expr_combined_log <- log2(expr_combined_norm + 1)
boxplot(expr_combined_log, main = "Post-normalization for both plexes", las = 2)

# Plex A
expr_A_norm <- normalizeBetweenArrays(expr_A_filt, method = "quantile")
expr_A_log <- log2(expr_A_norm + 1)
boxplot(expr_A_log, main = "Post-normalization for Plex A", las = 2)

# plex B
expr_B_norm <- normalizeBetweenArrays(expr_B_filt, method = "quantile")
expr_B_log <- log2(expr_B_norm + 1)
boxplot(expr_B_log, main = "Post-normalization for Plex B", las = 2)

```

## QC after normalization

After normalisation, the signal distributions are uniform, but there's a shoulder.

The correlation structure is retained as expected.

```{r, qc-post-normalization}

qc_long_norm_combined <- as.data.frame(expr_combined_log) %>%
  rownames_to_column("Protein") %>%
  pivot_longer(-Protein, names_to = "Sample", values_to = "Intensity") %>%
  filter(is.finite(Intensity))

ggplot(qc_long_norm_combined,
       aes(x = Intensity, color = Sample)) +
  geom_density(linewidth = 0.8) +
  theme_bw() +
  labs(title = "Post-normalization Intensity Distribution (Combined)",
       x = "Log2 Intensity",
       y = "Density")


cor_post_combined <- cor(expr_combined_log, use = "pairwise.complete.obs")

pheatmap(cor_post_combined,
         main = "Sample Correlation Heatmap (Post-normalization)")

```

## Impute missing values

Here we are using the median method for imputation of missing values.

```{r, impute1}

# Combine Plexes
expr_combined_imp <- expr_combined_log
for(i in 1:ncol(expr_combined_imp)){
  expr_combined_imp[is.na(expr_combined_imp[,i]), i] <-
    median(expr_combined_imp[,i], na.rm = TRUE)
}

# Plex A
expr_A_imp <- expr_A_log

for(i in 1:ncol(expr_A_imp)){
  expr_A_imp[is.na(expr_A_imp[,i]), i] <-
    median(expr_A_imp[,i], na.rm = TRUE)
}

# Plex B
expr_B_imp <- expr_B_log

for(i in 1:ncol(expr_B_imp)){
  expr_B_imp[is.na(expr_B_imp[,i]), i] <-
    median(expr_B_imp[,i], na.rm = TRUE)
}
sum(is.na(expr_combined_imp))
sum(is.na(expr_A_imp))
sum(is.na(expr_B_imp))

```

Next, we will removing Zero Variance Proteins because they cannot be statistically significant and may
cause other problems later on.

```{r,remove-zero-var}

expr_combined_imp <- expr_combined_imp[apply(expr_combined_imp, 1, var) != 0, ]
expr_A_imp <- expr_A_imp[apply(expr_A_imp, 1, var) != 0, ]
expr_B_imp <- expr_B_imp[apply(expr_B_imp, 1, var) != 0, ]

```

## Metadata Creation

```{r}
############################################
# METADATA CREATION (SINGLE SOURCE OF TRUTH)
############################################

# Create combined metadata from expression matrix
meta_combined <- data.frame(
  Sample = colnames(expr_combined_imp)
)

############################################
# Define Plex (based on known Plex A samples)
############################################

plexA_samples <- c(
  "19191","34629","34862","40431","40475","35038",
  "34866","41436","41383","40316","3746","102953"
)

meta_combined$Plex <- ifelse(meta_combined$Sample %in% plexA_samples,
                             "A", "B")

meta_combined$Plex <- factor(meta_combined$Plex)

############################################
# Define biological group (Case vs Control)
############################################

case_samples <- c(
  "19191","40797","34862","103376","40475","47442",
  "34866","43081","41383","113856","3746","118117"
)

meta_combined$Group <- ifelse(meta_combined$Sample %in% case_samples,
                              "Case", "Control")

meta_combined$Group <- factor(meta_combined$Group,
                              levels = c("Control","Case"))

############################################
# Final formatting
############################################

rownames(meta_combined) <- meta_combined$Sample

############################################
# Create Plex-specific metadata (CORRECT WAY)
############################################

meta_A <- meta_combined[meta_combined$Plex == "A", ]
meta_B <- meta_combined[meta_combined$Plex == "B", ]

############################################
# Align with expression matrices (CRITICAL)
############################################

meta_A <- meta_A[colnames(expr_A_imp), ]
meta_B <- meta_B[colnames(expr_B_imp), ]

############################################
# Keep only needed columns
############################################

meta_A <- meta_A[, c("Sample","Group")]
meta_B <- meta_B[, c("Sample","Group")]

############################################
# SANITY CHECKS (DO NOT SKIP)
############################################

# Alignment
all(rownames(meta_A) == colnames(expr_A_imp))
all(rownames(meta_B) == colnames(expr_B_imp))
all(rownames(meta_combined) == colnames(expr_combined_imp))

# Distribution checks
table(meta_combined$Plex, meta_combined$Group)
table(meta_A$Group)
table(meta_B$Group)
```

## PCA Before Batch Correction for combined plexes

```{r}
pca_pre <- prcomp(t(expr_combined_imp), center = TRUE, scale. = TRUE)

pca_pre_df <- data.frame(
  PC1 = pca_pre$x[,1],
  PC2 = pca_pre$x[,2],
  Sample = colnames(expr_combined_imp),
  Plex = meta_combined$Plex,
  Group = meta_combined$Group
)

ggplot(pca_pre_df, aes(x = PC1, y = PC2, color = Plex, shape = Group)) +
  geom_point(size = 4) +
  geom_text(aes(label = Sample), vjust = -0.5, size = 3) +
  theme_bw() +
  labs(title = "PCA Before Batch Correction",
       x = "PC1",
       y = "PC2")
```

## batch Correction

```{r}
expr_combined_corrected <- removeBatchEffect(
  expr_combined_imp,
  batch = meta_combined$Plex,
  design = model.matrix(~ Group, data = meta_combined)
)

# Quick check
head(expr_combined_corrected[,1:5])
# Dimensions unchanged
dim(expr_combined_corrected)
```

## PCA On Combined Plexes after removing Batch effect

```{r}
pca_post <- prcomp(t(expr_combined_corrected), center = TRUE, scale. = TRUE)

pca_post_df <- data.frame(
  PC1 = pca_post$x[,1],
  PC2 = pca_post$x[,2],
  Sample = colnames(expr_combined_corrected),
  Plex = meta_combined$Plex,
  Group = meta_combined$Group
)

ggplot(pca_post_df, aes(x = PC1, y = PC2, color = Group, shape = Plex)) +
  geom_point(size = 4) +
  geom_text(aes(label = Sample), vjust = -0.5, size = 3) +
  theme_bw() +
  labs(title = "PCA After Batch Correction",
       x = "PC1",
       y = "PC2")
```

## Differential Expression Analysis

### Combined plexes

```{r,de_combined}

design_combined <- model.matrix(~ Plex + Group, data = meta_combined)
fit_combined <- lmFit(expr_combined_imp, design_combined)
fit_combined <- eBayes(fit_combined)
results_combined <- topTable(
  fit_combined,
  coef = "GroupCase",
  number = Inf,
  sort.by = "P"
)
results_combined$Protein <- rownames(results_combined)
# Add gene annotation
results_combined <- merge(
  results_combined,
  protein_info_combined,
  by.x = "Protein",
  by.y = "protein",
  all.x = TRUE
)
# Add significance labels
results_combined$Significance <- "Not Significant"

results_combined$Significance[
  results_combined$adj.P.Val < 0.05 & results_combined$logFC > 1
] <- "Upregulated"
results_combined$Significance[
  results_combined$adj.P.Val < 0.05 & results_combined$logFC < -1
] <- "Downregulated"
# View

results_combined <- results_combined[order(results_combined$P.Value),]

head(results_combined) |>
    kbl(caption = "Top ranked proteins by p-value") |>
    kable_paper("hover", full_width = F)

```

## Separate plexes

Analyse plex A and plex B.

```{r,de_plex_A}

design_A <- model.matrix(~ Group, data = meta_A)
fit_A <- lmFit(expr_A_imp, design_A)
fit_A <- eBayes(fit_A)
results_A <- topTable(
  fit_A,
  coef = "GroupCase",
  number = Inf,
  sort.by = "P"
)
results_A$Protein <- rownames(results_A)
# Add gene annotation
results_A <- merge(
  results_A,
  protein_info_A,
  by.x = "Protein",
  by.y = "protein",
  all.x = TRUE
)

# Add significance labels
results_A$Significance <- "Not Significant"
results_A$Significance[
  results_A$adj.P.Val < 0.05 & results_A$logFC > 1
] <- "Upregulated"
results_A$Significance[
  results_A$adj.P.Val < 0.05 & results_A$logFC < -1
] <- "Downregulated"

# View
results_A <- results_A[order(results_A$P.Value),]

head(results_combined) |>
    kbl(caption = "Top ranked proteins by p-value in plex A") |>
    kable_paper("hover", full_width = F)

head(results_A)

```

Plex B.

```{r,de_plex_B}

design_B <- model.matrix(~ Group, data = meta_B)
fit_B <- lmFit(expr_B_imp, design_B)
fit_B <- eBayes(fit_B)
results_B <- topTable(
  fit_B,
  coef = "GroupCase",
  number = Inf,
  sort.by = "P"
)
results_B$Protein <- rownames(results_B)
# Add gene annotation
results_B <- merge(
  results_B,
  protein_info_B,
  by.x = "Protein",
  by.y = "protein",
  all.x = TRUE
)
# Add significance labels
results_B$Significance <- "Not Significant"
results_B$Significance[
  results_B$adj.P.Val < 0.05 & results_B$logFC > 1
] <- "Upregulated"
results_B$Significance[
  results_B$adj.P.Val < 0.05 & results_B$logFC < -1
] <- "Downregulated"

# View
results_B <- results_B[order(results_B$P.Value),]

head(results_combined) |>
    kbl(caption = "Top ranked proteins by p-value in plex B") |>
    kable_paper("hover", full_width = F)

head(results_B)


```

## Comparison of Plex A and B results

The top 50 up and down regulated proteins are selected from Plex A and B and these lists were compared.


```{r,compare_A_B}

plexA_up <- head(subset(results_A,logFC>0),50)[,1]
plexA_dn <- head(subset(results_A,logFC<0),50)[,1]
plexB_up <- head(subset(results_B,logFC>0),50)[,1]
plexB_dn <- head(subset(results_B,logFC<0),50)[,1]

v1 <- list("A up"=plexA_up, "A dn"=plexA_dn, "B up"=plexB_up, "B dn"=plexB_dn)

plot(euler(v1), quantities = TRUE)

intersect(plexA_up,plexB_up)

intersect(plexA_dn,plexB_dn)

intersect(plexA_up,plexB_dn)

intersect(plexB_up,plexA_dn)

```

## Ranking of proteins by logFC

```{r,rank_analysis}

results_A$score <- sign(results_A$logFC) * -log10(results_A$P.Value)
results_B$score <- sign(results_B$logFC) * -log10(results_B$P.Value)

results_m <- merge(results_A,results_B,by="Protein")

ranks_m <- results_m[,c("Protein","score.x","score.y")]
rownames(ranks_m) <- ranks_m$Protein ; ranks_m$Protein=NULL
colnames(ranks_m) <- c("Plex A","Plex B")

plot(ranks_m)

mylm <- lm(ranks_m$`Plex B` ~ ranks_m$`Plex A`)
abline(mylm,lwd=2,lty=2,col="red")
COR=signif(cor(ranks_m),3)[2,1]
COR
mtext(paste("Pearson correlation=",COR))

```

## Pathway Enrichment Analysis

Using the combined protein profiles.

Some of the pathways are known to be involved in CVD.

```{r,mitch1}

m <- mitch_import(results_combined,DEtype="limma",geneIDcol="gene")

reactome_sets <- gmt_import("ReactomePathways_2026-04-02.gmt")

mres <- mitch_calc(m,  reactome_sets, minsetsize = 5, cores=8)

head(mres$enrichment_result,10) |>
    kbl(caption = "Top ranked Reactome pathways by p-value") |>
    kable_paper("hover", full_width = F)

head(subset(mres$enrichment_result,s.dist>0),10) |>
    kbl(caption = "Top ranked upregulated Reactome pathways by p-value") |>
    kable_paper("hover", full_width = F)

head(subset(mres$enrichment_result,s.dist<0),10) |>
    kbl(caption = "Top ranked upregulated Reactome pathways by p-value") |>
    kable_paper("hover", full_width = F)

if ( file.exists("mitchreport1.html") ) { unlink("mitchreport1.html") }
mitch_report(mres,"mitchreport1.html")

```







## IG Filtering

```{r}
# plex A
ig_filter_A <- grepl("IG|KV|LV|IGH|IGK|IGL", rownames(expr_A_imp))
expr_A_clean <- expr_A_imp[!ig_filter_A, ]

# Plex B
ig_filter_B <- grepl("IG|KV|LV|IGH|IGK|IGL", rownames(expr_B_imp))
expr_B_clean <- expr_B_imp[!ig_filter_B, ]

# DE For Plex A
design_A <- model.matrix(~ Group, data = meta_A)

fit_A_clean <- lmFit(expr_A_clean, design_A)
fit_A_clean <- eBayes(fit_A_clean)

results_A_clean <- topTable(fit_A_clean, number = Inf)
results_A_clean$Protein <- rownames(results_A_clean)

# DE for Plex B
design_B <- model.matrix(~ Group, data = meta_B)

fit_B_clean <- lmFit(expr_B_clean, design_B)
fit_B_clean <- eBayes(fit_B_clean)

results_B_clean <- topTable(fit_B_clean, number = Inf)
results_B_clean$Protein <- rownames(results_B_clean)

head(results_A_clean)
head(results_B_clean)

```

## Session Info

```{r}

sessionInfo()

save.image("Analysis2.rdata")

```

